Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex
Draft of SC2 case study section

\begin{table}
\caption{Summary of the largest VCF Zarr columns for the Our Future Health
    genotype dataset, with chunk sizes of 1,000 variants and 
    10,000 samples and applying the BitRound(5) filter to the floating point 
    fields.
    For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-ofh-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
\bottomrule
\bottomrule
\end{tabular}
\end{table}

@@

>>>>

\begin{table}
\caption{Summary of the largest VCF Zarr columns for the Our Future Health
    genotype dataset, with chunk sizes of 1,000 variants and 
    10,000 samples and applying the BitRound(5) filter to the floating point 
    fields.
    For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-ofh-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

\subsection{Case study: SARS-Cov-2 alignment data}
The SARS-CoV-2 pandemic resulted in viral sequence data collection at 
an unprecedented scale, with over 15 million whole genomes incorporated into 
one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale 
overwhelmed classical phylogenetic methods, leading to the development
of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}. 
The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled 
set of consensus sequences for 4,471,579 samples, addressing a range of systematic 
errors in earlier collections. The \texttt{sc2ts} inference tool uses VCF-Zarr as its 
input format, and Zhan et al have made the Viridian dataset,
aligned to the reference using MAFFT~\citep{katoh2013mafft},
available for download~\citep{zhan2023towards}. This dataset has quite
different properties to those examined in previous case studies, and 
illustrates the flexibility of Zarr across a range of biological applications.

One of the unusual properties of this dataset is that whole genome alignments 
are stored, such that information on all 29903 bases is stored for every 
sample. Given the number of samples and the fact that nearly all bases
[TODO be concrete?] contain variation it is simply more convenient to 
store the entire alignment for each sample than to manage the complexity
of subsetting the sites. Similarly, the usual convention in VCF Zarr 
is that the \texttt{variant\_alleles} field contains the REF and ALT 
alleles as observed (such that the zeroth allele is always the reference).
In this case it is more convenient to encode the alleles as the original
bases and IUPAC uncertainty codes, such that the allele 0 is always A, 1 is C etc.
Even in this case with very high sequence diversity, uncertainty codes and 
storing data for the entire genome, compression rates are very high.

Based on the batch 1 release of Viridian, the Zarr dataset contains
3,960,704 sequences of length 29903 base pairs along with 31 metadata columns
encoding information about those samples. This full dataset is stored in a
single 345MiB Zip file, 241MiB of which is used by the alignment data. This
gives an average compression ratio of 468X.

Computational accessibility is excellent. For example, computing the number of
samples with missing or ambiguous data at each site requires around 3 minutes,
iterating over sample chunks sequentially (5 lines of Python).
Retrieving the full alignment for a given sample requires 163ms and 
getting the calls for all samples at a given site 209ms. [TODO confirm hardware
details]. The storage of metadata and genetic data in a single file also
makes many queries efficient and convenient. For example, computing the 
average number of deletions in samples with a Scorpio designation of Alpha
required x seconds and 5 lines of code using Dask.
